---
title: Exonerations in the U.S.
subtitle: Analyzing race, sentence severity, and time between conviction and exoneration
author: Project Marvelous Hitmontop <br> Caleb Chin, Khushi Patel, Alexa Prague, Derrick Chia, Cynthia Lei
date: May 5, 2023
format:
  revealjs:
    theme: simple
    smaller: true
editor: visual
execute:
  echo: false
  cache: true
---

## Topic and motivation

```{r}
#| label: load-packages
#| execute:
#|  echo: false

options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("gt")

library(tidyverse)
library(janitor)
library(tidymodels)
library(parsnip)
library(infer)
library(tidytext)
library(gt)


set.seed(123)
```

```{r}
#| label: data-load-clean
#| execute:
#| echo: false


us_exonerations <- read_csv("data/us_exonerations.csv")

most_severe <- c("Death", "Life without parole", "Life")


us_exonerations <- us_exonerations |>
  clean_names() |>
  rowwise() |>
  mutate(
    tags_vc = str_split(string = tags, pattern = ";#"),
    ars = if_else("A" %in% tags_vc, 1, 0),
    cdc = if_else("CDC" %in% tags_vc, 1, 0),
    ciu = if_else("CIU" %in% tags_vc, 1, 0),
    csh = if_else("CSH" %in% tags_vc, 1, 0),
    cv = if_else("CV" %in% tags_vc, 1, 0),
    fem = if_else("F" %in% tags_vc, 1, 0),
    fed = if_else("FED" %in% tags_vc, 1, 0),
    hom = if_else("H" %in% tags_vc, 1, 0),
    ji = if_else("JI" %in% tags_vc, 1, 0),
    m = if_else("M" %in% tags_vc, 1, 0),
    nc = if_else("NC" %in% tags_vc, 1, 0),
    p = if_else("P" %in% tags_vc, 1, 0),
    ph = if_else("PH" %in% tags_vc, 1, 0),
    sbs = if_else("SBS" %in% tags_vc, 1, 0),
    sa = if_else("SA" %in% tags_vc, 1, 0)
  ) |>
  ungroup() |>
  mutate(
    dna = if_else(is.na(dna), 0, 1),
    fc = if_else(is.na(fc), 0, 1),
    mwid = if_else(is.na(mwid), 0, 1),
    f_mfe = if_else(is.na(f_mfe), 0, 1),
    p_fa = if_else(is.na(p_fa), 0, 1),
    om = if_else(is.na(om), 0, 1),
    ild = if_else(is.na(ild), 0, 1),

    diff_conv_ex = exonerated - convicted,

    sentence_severity = if_else(sentence %in% most_severe, "Yes", "No"),
    sentence_severity = factor(sentence_severity, levels = c("Yes", "No"))
  ) |>
  select(-om_tags, -tags_vc, -list_addl_crimes_recode, -tags, -x)
```

-   The criminal justice system in the United States has been under scrutiny for its disproportionate impact on minority communities.

-   We want to investigate:

    -   how race affects the severity of the sentences exonerees receive

    -   how length of time between conviction and exoneration, severity of crime, and sentence are related

## Data introduction

- Observations represent all known U.S. exonerees between 1989 and 2023.
- Data was collected by The National Registry of Exonerations 
- The data fields consist of exoneree information such as name, race, and sex as well as information about their exoneration.
- Data has several subpopulations

## Highlights from EDA

::: {.r-stack}

```{r}
#| label: eda-summarystats
#| execute:
#| eval: TRUE
#| echo: FALSE
  
# proportion of exonerees by race
us_exonerations |>
  group_by(race) |>
  summarize(
    num = n()
  ) |>
  ungroup() |>
  mutate(
    prop = num / sum(num)
  ) |>
  gt()


```

::: {.fragment fragment-index=2}

```{r}
#| label: eda-race-exoneration-years
#| fig-height: 4

# box plot to show median length between conviction and exoneration by race
# for visualization purposes, we consider only 5 races, Asian, Hispanic, Black, White, and Native American
race_5 <- c('Asian', 'Hispanic', 'Black', 'White', 'Native American')

us_exonerations |>
  filter(race %in% race_5) |>
  ggplot(aes(x = race, y = diff_conv_ex, color = race)) +
  geom_boxplot(show.legend = FALSE) +
  scale_color_viridis_d() +
  labs(
    title = "Median years between conviction and exoneration by race",
    subtitle = "U.S. Exonerees, 1989 to 2023",
    x = "Exoneree Race",
    y = "Years between conviction and exoneration",
    caption = "Source: The National Registry of Exonerations"
  ) +
  theme_minimal()

```
:::
:::


Asian, Black, White, Hispanic, and Native American exonerees appear the most in the data set

::: {.fragment fragment-index=2}
Native American and Black exonerees often spend longer between conviction and exoneration than exonerees of other races.
:::


## Predicting sentence severity with race

::: r-stretch
```{r}
#| label: logi-reg-race-severity

race_5 <- c('Asian', 'Hispanic', 'Black', 'White', 'Native American')

fil_race5_exons <- us_exonerations |>
  filter(race %in% race_5) |>
  mutate(race = factor(race, levels = c("White", "Black", "Hispanic", "Asian", "Native American")))

log_fit_race_sent <- logistic_reg() |>
  fit(sentence_severity ~ race, data = fil_race5_exons)

race_pred <- tibble(race = c("White", "Black", "Hispanic", "Native American", "Asian"))
race_sent_predict <- bind_cols(race_pred, predict(log_fit_race_sent, new_data = race_pred, type="prob"))

race_sent_predict |>
  mutate(
    race = factor(race),
    race = fct_reorder(.f = race, .x = .pred_Yes)
  ) |>
  ggplot(aes(y = race, x = .pred_Yes, fill = race)) +
  geom_col(orientation = "y", show.legend = FALSE) + 
  expand_limits(x = c(0.0, 0.4)) +
  scale_fill_viridis_d(option = "C") +
  scale_x_continuous(labels=percent_format()) +
  labs(
    title = "Predicted Probability of Severe Sentence by Race",
    subtitle = "For U.S. Exonerees from 1989-2023, using logistic regression model",
    x = "Predicted probability of receiving death or life in prison sentence",
    y = "Race",
    caption = "Source: National Registry of Exonerations"
  ) +
  theme_minimal()

```
:::

::: fragment
Significance of results at 5% significance level: Hispanic Exonerees vs White Exonerees
:::

## Time falsely convicted, race, and sentence severity

::: r-stretch
```{r}
#| label: lin-reg-conv-sentsev-race

lin_fit_exons <- linear_reg() |>
  fit(diff_conv_ex ~ sentence_severity + race, data = fil_race5_exons)

val_race_sent <- expand(fil_race5_exons, race, sentence_severity)

pred_race_sent <- bind_cols(val_race_sent, predict(lin_fit_exons, new_data = val_race_sent))

pred_race_sent |>
  mutate(
    sentence_severity = as.factor(sentence_severity),
    sentence_severity = recode(sentence_severity, "Yes" = "Death or Life in Prison", "No" = "Other Sentence"),
    race = reorder_within(x = race, by = .pred, within = sentence_severity)
    ) |>
  ggplot(aes(x = race, y = .pred, fill = race)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(facets = vars(sentence_severity), scales = "free_x") +
  scale_x_reordered() +
  scale_fill_viridis_d(option = "C") +
  labs(
    title = "Predicting time falsely convicted with race and sentence severity",
    subtitle = "For U.S. Exonerees (1989-2023), using linear regression model",
    y = "Predicted number of years false convicted",
    x = "Exoneree Race",
    caption="Source: National Registry of Exonerees"
    
  ) +
  theme_minimal()
  

```
:::

::: {.fragment}
Significance of results at 5% significance level: Black Exonerees vs White Exonerees, Sentence Severity
:::

## Conclusions + future work

- Hispanic exonerees have a stat. significant difference in predicted probabiliy of getting severe sentence than White exonerees
- How about the rest of the exonerees?

::: {.fragment}
- Black exonerees have stat. significant difference compared to White exonerees in expected years between conviction and exoneration, holding sentence severity constant.
- What does this imply?
:::

::: {.fragment}
- Key Limitation: data only represents those who have already been found innocent, we cannot extrapolate to all falsely convicted people.
:::

::: {.fragment}
- Future work: 
  - Transforming sentence into numeric field to examine how sentence length (in years) is associated with race and/or other factors
:::
